Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2014-12-16 ms.tex
updated 3.1B
arg chaos

\label{sec:results}
\subsection{Properties near the stagnation radius}

@@

@@

Our above analysis neglects heat conduction near the stagnation radius, an assumption we can now check.
The importance of conductive heating can be estimated by comparing its
magnitude to that of wind heating, viz.~
\begin{align}
  \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}(r_{\rm s}) v_{\rm w}^{2}(1+\phi)}
\nonumber \\ &\sim {\rm min}
  \begin{cases}
  10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.8} v_{500}^{6.8} &  \text{unsaturated (core)}\\
 10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.5} v_{500}^{5.4} &  \text{unsaturated (cusp)}\\
  10 \phi/(2-\Gamma) & \text{saturated},
  \end{cases}
 \label{eq:rhors}
\end{align}
where in the second equality have made use of equations (\ref{eq:Tanalytic}), (\ref{eq:stag_simple}) and have approximated $\nabla^{2} \sim 1/r_{\rm st}^{2}$, $\nabla \sim 1/r_{\rm st}$.  Since the stagnation radius is usually well inside the break radius, the stellar density is approximated as (eq.~[\ref{eq:rhostar}])
\begin{eqnarray}
  \rho_{\star}(r_{\rm s}) &\simeq& \frac{M_{\bullet}(2-\Gamma)}{4\pi r_{\rm soi}^{3}}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-1-\Gamma} \nonumber \\
 &\approx& \begin{cases}
    9.4 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    1.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}
Even in the case of saturated heat flux, we see that for expected values $\phi < 1$, heat condution is at most moderately important at the stagnation radius, justifying its neglect in our analysis.

The SMBH accretion rate is simply given by the integrated mass loss
rate interior to the stagnation radius (equation~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    2.9 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    5.3 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
The Eddington ratio $\eddr$ is given by 
\begin{eqnarray}
\eddr&\approx&
  \begin{cases}
    1.3 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core} \\
    2.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
The density of gas at the stagnation radius $\rho(r_{\rm s})$ is more challenging to estimate.  We roughly estimate its value using equations (\ref{eq:stag_simple}) and (\ref{eq:mdot_analytic}) in conjunction with an alternative estimate of the mass accretion rate as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}(\rs) = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$, viz.
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{\tff(\rs)},
  \label{eq:mdot_gas}
\end{align}
from which one derives 
\begin{align}
  \rho(\rs)\simeq
  \begin{cases}
    5 \times 10^{-26} \Mbheight^{-0.2} v_{500}^{-0.8}  \eta_{0.02} \,
    \, {\rm g \, cm^{-3}}& \text{core}\\
    9 \times 10^{-26}  \Mbheight^{-0.5} v_{500}^{0.6}  \eta_{0.02} \,\, {\rm g \,cm^{-3}} & \text{cusp}
  \end{cases}
  \label{eq:rhors}
\end{align}

Note that combining equations~\eqref{eq:mdot_analytic}
and~\eqref{eq:mdot_gas} we can rewrite the $\rho(\rs)$

\begin{align}
\rho(\rs)=\frac{3 q(\rs) \tff(\rs)}{2-\Gamma}.
\label{eq:rhors2}
\end{align}

\subsection{Comparison with Bondi Accretion}
The spherical accretion rate onto a point source from a surrounding
medium of fixed sound speed $c_{\rm s} = (kT/\mu m_p)^{1/2}$ and
density is given by (\citealt{Bondi52})
\begin{align}
  \dot{M}_{\rm B} =4\pi \lambda r_b^2 \rho(r_b) v_{\rm ff}(r_b),
\label{eq:bondi}
\end{align}
where $r_b \equiv GM/c_{s}^{2}$ is the ``Bondi radius", $v_{\rm ff} = (GM_{\bullet}/r)^{1/2}$ and $\lambda$ is a parameter of order unity.

Rewriting equation (\ref{eq:mdot_gas}) as
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{\tff} =\frac{4
    \pi}{3} \rs^2 \rho(\rs) v_{\rm ff}(\rs)
\end{align}
we note the qualitative similarity between this expression and the Bondi formula (equation~[\ref{eq:bondi}]) with the exception that $\rb$ is replaced by $\rs$.  Using equation (\ref{eq:stag_simple})
\begin{align}
  \rs\approx\frac{7}{2}\frac{G \Mbh}{\vwO^2} \approx \frac{14}{3}\frac{GM_{\bullet}}{c_{s}^{2}},
  \label{eq:rs_simple}
\end{align}
where in the second line we have used the fact that $v_{w}^{2}/2 =
c_s^2/(\gamma-1)$ at $\rs$.  Thus, provided that a stagnation radius
exists which is interior to the sphere of influence radius, the
accretion rate predicted by our model is quite close to $\dot{M}_{\rm
  B}$.
 %({\bf BDM: Check - as this depends on
 % whether eq 10 is correct})

%% AG-Apparently even in the low heating case, it is still possible to
%% an outflow (from the analytic results above the gas density must be
%% quite steep in this case). However, at low heating we will probably
%% become subject to a cooling instablity.  AG-I believe that in the
%% case of core galaxy, the negative root of the quadratic should be
%% taken...

\subsection{Numerical Solutions}

Numerical solutions are calculated for each galaxy for three different values of the stellar wind heating parameter $v_{w,0}$ = 300, 600, 1200 km s$^{-1}$.

@@

\subsection{Cooling}
\label{sec:cooling}
Our models neglect radiative cooling, an assumption we now check.  At high temperatures $T\gsim 2\times 10^7$ K gas cools through free-free emission, while at lower temperatures it cools predominantly through metal lines.  To assess the importance of radiative cooling in our solutions we compare the heating rate $H=q \vw^2/2$ to the nominal cooling rate
\begin{align}
  \dot{q}_{\rm cool} =\Lambda(T) n^2,
\label{eq:cooling}
\end{align}
where $n = \rho/\mu m_p$ and we approximate the cooling function $\Lambda(T)$ assuming solar metallicity gas using piece-wise power law (Fig. 34.1 in \cite{Draine:2011a}):
\begin{align}
\Lambda(T)\simeq
  \begin{cases}
    2.3 \times 10^{-24} \left(T/10^6 \text{K}\right)^{0.5} $erg cm$^3 $s$^{-1}&, T \gsim 2 \times 10^7 \text{K} \\
    1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}&, T \lsim 2 \times 10^7 \text{K}     
  \end{cases}
  \label{eq:cooling2}
\end{align}

Radiative cooling can have an important on even the qualitative features of the flow, because initially hydrostatic gas that cools at constant pressure $P \propto nT$ according to $\dot{q}_{\rm cool} \propto T^{-2.7}$ (eq.~[\ref{eq:cooling}]) is thermally unstable.  Radiative cooling can be stabilized, however, by the other source of (stable) wind cooling, given by the last term $\propto T$ in the entropy equation (eq.~[\ref{eq:dsdt}]), which at the stagnation radius equals $q v_{\rm w}^{2}/2$.

An important ratio controlling the importance of thermal stability is thus
\begin{align}
\left.\frac{q \vw^2/2}{\dot{q}_{\rm cool}}\right|_{\rm r{s}} \simeq
  \begin{cases}
   590 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   210 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhostarrs}) and
(\ref{eq:rhors}).  We thus see that we require $v_{\rm w} \gtrsim $
300 km s$^{-1}$ for stability, with a weak sensitivity on $\eta$ and
$M_{\bullet}$.  {\bf AG:I flipped the ratio to be consistent with the
  ratio of timescales given below (NB here we are comparing heating
  and cooling rates as opposed to timescales). Modulo factors of order
  unity I find no inconsistency. We should probably only keep one of
  these ratios. Note I have adopted the new convention of Gamma=0.8 is
  a cusp galaxy and Gamma=0.1 is a core galaxy.}

At the stagnation $\tcool/\tff$ is equal to the ratio of the heating to
the cooling rate to factors of order unity as shown below

\begin{align}
\left. \frac{\tcool}{\tff}\right|_{\rs}&=\left. \frac{3}{5} \frac{\rho 
  \vw^2/2}{\dot{q}_{\rm cool}}\right|_{\rs}\\
&=\frac{9}{5} \left. \frac{q \vw^2/2}{\dot{q}_{\rm cool}}\right|_{\rs}.
\end{align}

In the first line we have used equation~\eqref{eq:Tanalytic} and in
the second line we have used equation~\eqref{eq:rhors2}.

>>>>

\label{sec:results}
\subsection{Analytic Results}

@@

\subsubsection{Properties at the Stagnation Radius}

@@

\subsubsection{SMBH Accretion Rate}

The SMBH accretion rate is simply given by the integrated mass loss
rate interior to the stagnation radius (equation~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    2.9 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    5.3 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
The Eddington ratio $\eddr$ is given by 
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    1.3 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core} \\
    2.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}

The density of gas at the stagnation radius $\rho(r_{\rm s})$ is more challenging to estimate.  We roughly estimate its value using equations (\ref{eq:stag_simple}) and (\ref{eq:mdot_analytic}) in conjunction with an alternative estimate of the mass accretion rate as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}(\rs) = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$, viz.
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{\tff(\rs)},
  \label{eq:mdot_gas}
\end{align}
from which one derives 
\begin{align}
  \rho(\rs)\simeq
  \begin{cases}
    5 \times 10^{-26} \Mbheight^{-0.2} v_{500}^{-0.8}  \eta_{0.02} \,
    \, {\rm g \, cm^{-3}}& \text{core}\\
    9 \times 10^{-26}  \Mbheight^{-0.5} v_{500}^{0.6}  \eta_{0.02} \,\, {\rm g \,cm^{-3}} & \text{cusp}
  \end{cases}
  \label{eq:rhors}
\end{align}
Note that by combining equations~\eqref{eq:mdot_analytic}
and~\eqref{eq:mdot_gas}, one can also write
\begin{align}
\rho(\rs)=\frac{3 q(\rs) \tff(\rs)}{2-\Gamma}.
\label{eq:rhors2}
\end{align}

Our expression for the mass accretion rate (eq.~[\ref{eq:mdot_analytic}]) is usefully compared to the traditional ``Bondi" accretion rate onto a point source from a surrounding medium of fixed sound speed $c_{\rm s} = (kT/\mu m_p)^{1/2}$ and
density (\citealt{Bondi52}):
\begin{align}
  \dot{M}_{\rm B} =4\pi \lambda r_b^2 \rho(r_b) v_{\rm ff}(r_b),
\label{eq:bondi}
\end{align}
where $r_b \equiv GM/c_{s}^{2}$ is the Bondi radius, $v_{\rm ff} = (GM_{\bullet}/r)^{1/2}$ and $\lambda$ is a parameter of order unity.  Rewriting equation (\ref{eq:mdot_gas}) as $\dot{M}\simeq 4\pi \rs^2 \rho(\rs) v_{\rm ff}(\rs)/3$,
we note a similarity Bondi formula (eq.~[\ref{eq:bondi}]) with the exception that $\rb$ is replaced by $\rs$.  In fact, when the stagnation radius resides interior to the sphere of influence we have (eq.~[\ref{eq:stag_simple}])
\begin{align}
  \rs\approx\frac{7}{2}\frac{G \Mbh}{\vwO^2} \approx \frac{14}{3}\frac{GM_{\bullet}}{c_{s}^{2}} = \frac{14}{3}r_{\rm B},
  \label{eq:rs_simple}
\end{align}
where in the second line we have used equation (\ref{eq:Tanalytic}).

\subsubsection{Importance of Heat Conduction}

Our analytic results neglect the effects of heat conduction near the stagnation radius, an assumption we now check.  The importance of conductive heating can be estimated by comparing its magnitude to that of ``wind" heating, viz.~
\begin{align}
  \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}(r_{\rm s}) v_{\rm w}^{2}(1+\phi)}
\nonumber \\ &\sim {\rm min}
  \begin{cases}
  10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.8} v_{500}^{6.8} &  \text{unsaturated (core)}\\
 10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.5} v_{500}^{5.4} &  \text{unsaturated (cusp)}\\
  10 \phi/(2-\Gamma) & \text{saturated},
  \end{cases}
 \label{eq:rhors}
\end{align}
where the second equality makes use of equations (\ref{eq:Tanalytic}), (\ref{eq:stag_simple}), and we have approximated $\nabla^{2} \sim 1/r_{\rm st}^{2}$, $\nabla \sim 1/r_{\rm st}$.  Because the stagnation radius is usually well inside the break radius, the stellar density is approximated as (eq.~[\ref{eq:rhostar}])
\begin{eqnarray}
  \rho_{\star}(r_{\rm s}) &\simeq& \frac{M_{\bullet}(2-\Gamma)}{4\pi r_{\rm soi}^{3}}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-1-\Gamma} \nonumber \\
 &\approx& \begin{cases}
    9.4 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    1.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}
Even in the saturated heat flux case, heat conduction is at most moderately important for expected values of the saturation parameter $\phi < 1$, justifying its neglect in our analysis.

\subsubsection{Thermal Instability}

Radiative cooling, if important, can qualitative alter key features of the flow.  Initially hydrostatic gas that cools at constant pressure $P \propto nT$ according to $\dot{q}_{\rm cool} \propto T^{-2.7}$ (eq.~[\ref{eq:cooling}] below) is thermally unstable.  Thermal instability leads to higher black hole accretion rates and enhanced star formation which, when coupled by feed-back processes, can lead to limit cycle behavior, inconsistent with our assumption of a a steady-state flow (e.g. Ciotti refs).

Radiative cooling can, however, be stabilized by the other source of (stable) wind cooling, such as the last term $\propto c_{s}^{2} \propto T$ in the entropy equation (eq.~[\ref{eq:dsdt}]), which at the stagnation radius equals $q v_{\rm w}^{2}/2$.  To assess the conditions for thermal stability we compare the nominal rate of wind heating to the rate of radiative cooling $\dot{q}_{\rm cool} =\Lambda(T) n^2$ at the stagnation radius,
\begin{align}
\left.\frac{q \vw^2/2}{\dot{q}_{\rm cool}}\right|_{\rm r{s}} \simeq
  \begin{cases}
   590 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   210 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhostarrs}) and
(\ref{eq:rhors}).  Here $n = \rho/\mu m_p$ and we have approximated the cooling function $\Lambda(T)$ assuming solar metallicity gas using piece-wise power law (\citealt{Draine:2011a}, Fig. 34.1):
\begin{align}
\Lambda(T)\simeq
  \begin{cases}
    2.3 \times 10^{-24} \left(T/10^6 \text{K}\right)^{0.5} $erg cm$^3 $s$^{-1}&, T > 2 \times 10^7 \text{K} \\
    1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}&, T < 2 \times 10^7 \text{K}     
  \end{cases}
  \label{eq:cooling2}
\end{align}
Note that at the stagnation radius equation (\ref{eq:cooling2}) is equivalent to the ratio of the cooling to free-fall timescale, i.e.
\begin{align}
\left. \frac{\tcool}{\tff}\right|_{\rs}&= \frac{9}{5} \left. \frac{q \vw^2/2}{\dot{q}_{\rm cool}}\right|_{\rs},
\end{align}
where we have used equations~(\eqref{eq:Tanalytic}) and (\eqref{eq:rhors2}).

There thus exists a critical minimum wind velocity required for thermal stability at the stagnation radius
\begin{align}
v_{\rm crit} \simeq
  \begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
In $\S\ref{sec:numerical}$ we will show that $v_{w} \gtrsim v_{\rm crit} \sim 200-300$ km s$^{-1}$ is generally a condition for stability at all radii, not just at the stagnation radius.

\subsection{Numerical Results}
\label{sec:numerical}

This section describes the results of our numerical calculations and tests the accuracy of the analytic expressions provided in the previous section.  Numerical solutions are calculated for each galaxy for three different values of the stellar wind heating parameter $v_{w,0}$ = 300, 600, 1200 km s$^{-1}$.

@@

\subsection{Cooling}
\label{sec:cooling}
